---
title: "ChIP-seq annotation & visualization with ChIPseeker"
author: "Raja Khan"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    highlight: tango
    df_print: paged
    fig_caption: true
    self_contained: true
---

```{r setup, include=TRUE}
library(ChIPseeker)
library(GenomicRanges)
library(GenomicFeatures)
library(AnnotationDbi)
library(clusterProfiler)
library(ggplot2)

setwd("/Volumes/CrucialX6/human/")

knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE,
                      fig.width = 8, fig.height = 5, dpi = 600)
```

# Parameters

```{r params}
peakdir  <- "19.peaks_chr11_12/"
gtf      <- "gencode.v49.annotation.gtf"

plot_theme <-  theme(text = element_text(size = 12, colour = "black", family = "Helvetica"),
        axis.text.y = element_text(size = 12, colour = "black", family = "Helvetica"),
        axis.text.x = element_text(size = 12, colour = "black", family = "Helvetica"),
        axis.title = element_text(size = 12, colour = "black", family = "Helvetica"),
        axis.line = element_line(linewidth = 0.3528), axis.ticks = element_line(linewidth = 0.3528),
        legend.text  = element_text(size = 12, family = "Helvetica", colour = "black"))
```

# Build TxDb from GTF

```{r txdb}
txdb <- makeTxDbFromGFF(gtf, format = "gtf")
tx <- transcripts(txdb)
length(tx)
```

# Get peak files and sample names

```{r peaks}
samplefiles <- list.files(peakdir, full.names = TRUE)
sample_names = basename(samplefiles)

sample_names <- sub("\\.(narrowPeak|broadPeak|bed|bedgraph|bam)(\\.gz)?$", "", 
                    sample_names, ignore.case = TRUE)
sample_names <- sub("\\.chr11_12_peaks$", "", sample_names, ignore.case = TRUE)

files_for_peakprof <- setNames(as.list(samplefiles), sample_names)

head(sample_names)
```

# Single Peak analysis

```{r}
ceb      <- files_for_peakprof[["ceb_ENCFF327JFG"]]
H3K27me3 <- files_for_peakprof[["H3K27me3_ENCFF164ALR"]]
H3K9ac   <- files_for_peakprof[["H3K9ac_ENCFF193NPE"]]
```

### Figure: TSS heatmap for CEB peaks  
This heatmap shows CEB read density ±3 kb around transcription start sites, highlighting promoter-centered binding patterns.

```{r}
peak_Profile_Heatmap(
  peak = ceb,
  upstream = 3000,
  downstream = 3000,
  by = "gene",
  type = "start_site",
  TxDb = txdb,
  nbin = 800
) 
```

### TSS heatmap for H3K27me3 peaks  
This heatmap visualizes how the repressive H3K27me3 mark distributes relative to TSS positions.

```{r}
peak_Profile_Heatmap(
  peak = H3K27me3,
  upstream = 3000,
  downstream = 3000,
  by = "gene",
  type = "start_site",
  TxDb = txdb,
  nbin = 800
)
```

### TSS heatmap for H3K9ac peaks  
This plot shows enrichment of the active histone mark H3K9ac near TSS regions.

```{r}
peak_Profile_Heatmap(
  peak = H3K9ac,
  upstream = 3000,
  downstream = 3000,
  by = "gene",
  type = "start_site",
  TxDb = txdb,
  nbin = 800
) 
```

# Promoter-centric tag matrix and average profile

### Average promoter profiles across samples  
This line plot shows averaged tag density across promoter windows for each dataset.

```{r tagmatrix}
promoter <- getPromoters(TxDb = txdb, upstream = 3000, downstream = 3000)

tagMatrixList <- setNames(
  lapply(samplefiles, getTagMatrix, windows = promoter),
  sample_names
)

p1 <- plotAvgProf(tagMatrixList, xlim = c(-3000, 3000)) + plot_theme

p1
```

# Peak body profiles (plotPeakProf2)

###  Normalized gene-body binding profiles  
This figure shows how signals distribute across scaled gene bodies with 20% upstream/downstream extension.

```{r peakprof}
plotPeakProf2(files_for_peakprof,
              upstream = rel(0.2), downstream = rel(0.2), by = "gene", type = "body",
              TxDb = txdb, nbin = 800) + plot_theme
```

# Annotate peaks and genomic distribution plots

### Genomic feature distribution for CEB peaks  
This barplot summarizes the fraction of CEB peaks in promoters, introns, exons, and intergenic space.

```{r}
peakAnno.ceb <- annotatePeak(
  ceb,
  tssRegion = c(-3000, 3000),
  TxDb = txdb
)

peakAnno.ceb
plotAnnoBar(peakAnno.ceb) + plot_theme
```

###  Genomic feature distribution for H3K27me3 peaks  
This annotation plot shows how H3K27me3 peaks distribute across genomic elements.

```{r}
peakAnno.H3K27me3 <- annotatePeak(
  H3K27me3,
  tssRegion = c(-3000, 3000),
  TxDb = txdb
)

peakAnno.H3K27me3
plotAnnoBar(peakAnno.H3K27me3) + plot_theme
```

### Genomic feature distribution for H3K9ac peaks  
This plot displays the annotation categories for H3K9ac peaks, a mark associated with active chromatin states.

```{r}
peakAnno.H3K9ac <- annotatePeak(
  H3K9ac,
  tssRegion = c(-3000, 3000),
  TxDb = txdb
)

peakAnno.H3K9ac
plotAnnoBar(peakAnno.H3K9ac) + plot_theme
```

### Combined annotation across all samples  
This multi-sample barplot compares how each dataset distributes across promoters, introns, exons, and intergenic regions.

```{r}
peakAnnoList <- lapply(samplefiles, annotatePeak, TxDb = txdb,
                       tssRegion = c(-3000, 3000), verbose = FALSE)
names(peakAnnoList) <- sample_names

plotAnnoBar(peakAnnoList) + plot_theme
```

### Distance-to-TSS distributions across samples  
This figure visualizes how the peaks of each dataset are distributed relative to TSS positions.

```{r}
plotDistToTSS(peakAnnoList) + plot_theme
```

# Notes and troubleshooting

- Ensure peak files are valid BED-like formats.
- Check sample naming if plots appear mislabeled.
- Use named lists for functions requiring sample identities.

# Session info

```{r session}
sessionInfo()
```
